---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
# Use setwd() to change the working directory to where activity.csv has been extracted from activity.zip.

activity <- read.csv( "activity.csv" )


```

## What is mean total number of steps taken per day?

```{r}
# create a subset free of missing data
activity_sans_na <- subset( activity, ! is.na (steps) )

# group the steps by date
total_steps_by_date <- aggregate( steps ~ date, activity_sans_na, sum )
names( total_steps_by_date ) <- c( "date", "total_steps" )

head( total_steps_by_date, n = 5 )
tail( total_steps_by_date, n = 5 )
```

```{r}
# plot the result
hist( total_steps_by_date$total_steps, xlab = "", main = "Total Steps Per Day" )

# mean of total number of steps taken per day
mean( total_steps_by_date$total_steps )

# median of total number of steps taken per day
median( total_steps_by_date$total_steps )
```

## What is the average daily activity pattern?

```{r}
# Calculate the averages
average_steps_every_5_minutes <- aggregate( steps ~ interval, activity_sans_na, mean )

# Plot the result
plot( type = "l", average_steps_every_5_minutes, xlab = "24 Hours in 5-minute Interval", ylab = "Average Steps Taken", main = "Average Daily Activities" )

# Find the interval that has the maximum number of steps
average_steps_every_5_minutes [ average_steps_every_5_minutes$steps == max( average_steps_every_5_minutes$steps), ]
```

It appears 08:35am is the most active interval during the day.

## Inputing missing values

Analyze the original dataset for missing data.

```{r}
summary( activity )

summary( activity$steps )

```
It looks like only the steps column contains missing values and there are 2304 observations/rows with NA.

Here's a sample of these observations:
```{r}
activity_with_na <- subset( activity, is.na (steps) )

head( activity_with_na )
tail( activity_with_na )
```

The strategy for filling missing values is to use the mean (average steps) for the corresponding interval.

```{r}
activity_na_replaced <- transform( activity, steps = ifelse( is.na (steps), average_steps_every_5_minutes[ average_steps_every_5_minutes$interval == interval, ]$steps, steps) )

```

Here is the histogram of this new version:
```{r}
# group the steps by date
total_steps_by_date <- aggregate( steps ~ date, activity_na_replaced, sum )
names( total_steps_by_date ) <- c( "date", "total_steps" )

# plot the result
hist( total_steps_by_date$total_steps, xlab = "", main = "Total Steps Per Day (NAs replaced)" )

# mean of total number of steps taken per day
mean( total_steps_by_date$total_steps )

# median of total number of steps taken per day
median( total_steps_by_date$total_steps )
```


## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_na_replaced$day <- weekdays( as.Date (activity_na_replaced$date) )

week_days <- c ("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

activity_na_replaced[ ! activity_na_replaced$day %in% week_days, "day"] <- "Weekend" 
activity_na_replaced[ activity_na_replaced$day %in% week_days, "day"] <- "Weekday" 

activity_na_replaced$day <- as.factor( activity_na_replaced$day )

summary( activity_na_replaced$day )

#head ( activity_na_replaced )
```

Here is the plot comparing the steps of weekdays and those on the weekends:
```{r}

activity_na_replaced_compare <- aggregate( activity_na_replaced$steps ~ activity_na_replaced$interval + activity_na_replaced$day, activity_na_replaced, mean )

names( activity_na_replaced_compare ) <- c("interval", "is_weekend", "average_steps")

# plot it
library( lattice )
par( mfrow = c(2,1) )

xyplot( average_steps ~ interval | is_weekend, data = activity_na_replaced_compare, type = "l", aspect = 2/5, ylab = "Number of steps", main = "Average Daily Pattern" )
```
